---
title: "Review of Regression"
author: "Andrew Zieffler"
date: "September 2, 2014"
theme: readable
output:
  ioslides_presentation:
    highlight: tango
    smaller: yes
font-import: http://fonts.googleapis.com/css?family=EB+Garamond
font-family: EB+Garamond  
font-size: 10pt
---

## Fan Cost Index

```{r}
nfl = read.csv(file = "~/Documents/EPsy-8252/data/FCI-NFL-2014.csv")
head(nfl)
```


## Meta data

```{r}
meta = read.csv(file = "~/Documents/EPsy-8252/data/NFL-Meta-Data.csv")
head(meta)
```

## Merge meta data into nfl data frame

```{r}
nfl = merge(nfl, meta, by = "team")
head(nfl, 4)
```

## Create age of stadium variable

```{r}
nfl$ageStadium = 2014 - nfl$yearOpened
head(nfl, 4)
```

## Examine outcome: Density plot

```{r,message=FALSE,fig.width=4,fig.height=4}
library(sm)
sm.density(nfl$fci)
```

## Log of outcome

```{r,message=FALSE,fig.width=4,fig.height=4}
nfl$Lfci = log(nfl$fci)
sm.density(nfl$Lfci)
```

## Examine outcome: Summaries

```{r,message=FALSE, warning=FALSE}
library(psych)
describe(nfl$Lfci)
```

## Lfci vs. ageStadium

```{r,message=FALSE,fig.width=4,fig.height=4,warning=FALSE,tidy=FALSE}
library(ggplot2)
ggplot(data = nfl, aes(x = ageStadium, y = Lfci)) +
  geom_point() +
	#geom_text(aes(label = team)) +
	geom_smooth(se = FALSE) +
	theme_bw()
```

## Quadratic outcome

```{r}
nfl$ageStadiumQ = nfl$ageStadium ^ 2
coef(summary(lm(Lfci ~ ageStadium + ageStadiumQ, data = nfl)))
```

## Scatterplot matrix

```{r,message=FALSE, warning=FALSE,tidy=FALSE}
library(car)
scatterplotMatrix(nfl[c("Lfci", "ageStadium", "ageStadiumQ", "pts2013",
    "ptsAgainst2013", "winPct2013", "coachYrswTeam", "newCoach", 
    "playoffs2013", "playoffWins2013", "lat", "lon", "capacity")])
```

## Correlation matrix

```{r}
corMat = cor(nfl[c("Lfci", "ageStadium", "ageStadiumQ", "pts2013", 
    "ptsAgainst2013", "winPct2013", "coachYrswTeam", "newCoach", 
    "playoffs2013", "playoffWins2013", "lat", "lon", "capacity")])
corMat
```


## Partial correlation matrix

```{r}
# Needs psych library - remove effects of age of stadium (L and Q)
partial.r(corMat, c(1, 4:13), c(2, 3))
```


## Examine coachYrswTeam predictor

```{r,fig.width=4,fig.height=4}
# Needs sm library
sm.density(nfl$coachYrswTeam)
```

## Lfci vs. coachYrswTeam 

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
ggplot(data = nfl, aes(x = coachYrswTeam, y = Lfci)) +
  geom_point() +
	#geom_text(aes(label = team)) +
	geom_smooth(se = FALSE) +
	theme_bw()
```

## Lfci vs. LcoachYrswTeam 

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
nfl$LcoachYrswTeam = log(nfl$coachYrswTeam + 1)
ggplot(data = nfl, aes(x = LcoachYrswTeam, y = Lfci)) +
  geom_point() +
	#geom_text(aes(label = team)) +
	geom_smooth(se = FALSE) +
	theme_bw()
```

## Regress Lfci on age of stadium and LcoachYrswTeam

```{r}
coef(summary(lm(Lfci ~ ageStadium + I(ageStadium ^ 2) + LcoachYrswTeam, data = nfl)))
```

## Partial correlation matrix: Part II

```{r}
corMat = cor(nfl[c("Lfci", "ageStadium", "ageStadiumQ", "pts2013", 
    "ptsAgainst2013", "winPct2013", "LcoachYrswTeam", "newCoach", 
    "playoffs2013", "playoffWins2013", "lat", "lon", "capacity")])

#remove effects of age of stadium (L and Q) and coachYrswTeam
partial.r(corMat, c(1, 4:6, 8:13), c(2, 3, 7)) 
```

## Examine capacity predictor

```{r,fig.width=4,fig.height=4}
# Needs sm library
sm.density(nfl$capacity)
```

## Lfci vs. capacity 

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
ggplot(data = nfl, aes(x = capacity, y = Lfci)) +
  geom_point() +
	#geom_text(aes(label = team)) +
	geom_smooth(se = FALSE) +
	theme_bw()
```

## Regress Lfci on age of stadium, LcoachYrswTeam, and capacity

```{r,tidy=FALSE}
coef(summary(lm(Lfci ~ ageStadium + I(ageStadium ^ 2) + LcoachYrswTeam +
  capacity, data = nfl)))
```

## Obtain residuals

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
lm.a = lm(fci ~ ageStadium + I(ageStadium ^ 2) + LcoachYrswTeam, data = nfl)
out.a = fortify(lm.a)
head(out.a)
```

## Evaluate residuals (normality)

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
sm.density(out.a$.stdresid, model = "normal")
```

## Evaluate residuals (linearity, homoscedasticity)

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
ggplot(data = out.a, aes(x = .fitted, y = .stdresid)) +
  geom_point(size = 3) +
	geom_hline(yintercept = 0) +
	theme_bw()
```

## Identify outlier

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
out.a$team = nfl$team
ggplot(data = out.a, aes(x = .fitted, y = .stdresid)) +
  geom_text(aes(label = team), size = 3) +
  geom_hline(yintercept = 0) +
  theme_bw()
```

## Identify outlier by row number for easier removal

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
ggplot(data = out.a, aes(x = .fitted, y = .stdresid)) +
  geom_text(aes(label = rownames(out.a)), size = 3) +
  geom_hline(yintercept = 0) +
  theme_bw()	
```

## Remove Washington (row 32)

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
lm.b = lm(fci ~ ageStadium + I(ageStadium ^ 2) + LcoachYrswTeam, data = nfl, subset = -c(32))
out.b = fortify(lm.b)
out.b$team = nfl$team[-c(32)] 
```

## Evaluate residuals (normality)

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
sm.density(out.b$.stdresid, model = "normal")
```

## Evaluate residuals (linearity, homoscedasticity)

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
ggplot(data = out.b, aes(x = .fitted, y = .stdresid)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0) +
  theme_bw()
```

## Prepare data set to plot Model B

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
nfl2 = expand.grid(
  ageStadium = 0:90,
  LcoachYrswTeam = c(0, 1.098612, 1.791759, 2.397895) # O, 2, and 5 years with team
  )

# Get predicted values for fci
preds = predict(lm.b, newdata = nfl2)
head(preds)
```

## Add predicted values to nfl2 data frame

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
nfl2 = cbind(nfl2, preds)
head(nfl2)
```

## Coerce LcoachYrswTeam into a factor for plotting

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
nfl2$LcoachYrswTeam = factor(nfl2$LcoachYrswTeam,
  levels = c(0, 1.098612, 1.791759, 2.397895),
  labels = c("0 years", "2 years", "5 years", "10 years")
	)
```

## Final plot 1

```{r,fig.width=6,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
ggplot(data = nfl2, aes(x = ageStadium, y = preds, group = LcoachYrswTeam)) +
    geom_line(aes(color = LcoachYrswTeam)) +
    xlab("Age of Stadium") +
    ylab("Predicted Cost") +
    theme_bw() +
    scale_color_brewer(name = "Time with team", palette = "Set2")
```

## Prepare data set to plot Model B for average LcoachYrswTeam

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
nfl2 = expand.grid(
  ageStadium = 0:90,
  LcoachYrswTeam = 1.136643
  )

# Get predicted values for fci
preds = predict(lm.b, newdata = nfl2)

# Add predicted values to data frame
nfl2 = cbind(nfl2, preds)
```

## Final plot 2

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
ggplot(data = nfl2, aes(x = ageStadium, y = preds)) +
    geom_line() +
    xlab("Age of Stadium") +
    ylab("Predicted Cost") +
    theme_bw()
```

## How much will it cost to go to a Vikings game in 2016?

```{r,fig.width=4,fig.height=4,tidy=FALSE,warning=FALSE,message=FALSE}
vikes = data.frame(
  ageStadium = 0,
  LcoachYrswTeam = 1.098612 # coach of 2 years
  )

# Get predicted values for fci
predict(lm.b, newdata = vikes, interval = "prediction")
```

